Initial Analysis

### 2019-10-23 Initial_Analysis.Rmd
Cleaned the script

1. Determine which countries have Neighbouring EEZs  
  1.1. For this we used the Marine regions EEZ maps (We have to change this because it does not discriminate by oceans and we need this) and a routine to identify Neighbouring polygons,
  1.2. Then we match the DBEM spatial information (e.g. INDEX within EEZ and coordinates) with the Neighbouring data
2. Overlap distribution maps with Neighbours data  
  2.1. For each species we merged the DBEM_Neighbours data with their known distribution (Gab's data)  
3. Determine transboundary if:  
  3.1. Species is present in both Neighbouring EEZs by INDEX

Alternatively, we could create a "confidence level" based on a ratio, i.e;
- How many grid cells of both the EEZs is the species present on. A species that covers all cells would have 100% but a species that covers, lets say, all of EEZ1 but only half of EEZ2 would be 75%.  
- Another option is to have the index based on the species distribution. That is to say, what proportion of the total species distribution actually falls within both EEZs....

## Data

### Determine species list

@@

## Estimating Neighbourds

### Functions `EstNeighbours`

@@

  }else{
  # Merge them with SAU data
  suppressWarnings(
    # warning for characters in name, no problem
  DBEM_Neighbours_Data <- Result_table %>% 
     left_join(EEZIDs_List, 
              by = "Name") %>% # Asign EEZ id to each country
    left_join(EEZ_CellID, 
              by = "EEZID") %>% # Asign EEZ id to each country
    left_join(Coor,
              by = "INDEX") %>%  # Inclu
    rename(Country_Territory = Name)

@@

```{r EEZ_Neighbourds_Control_Pannel, eval = F, echo = F}

#### Load Data Needed ####

# Spatial analysis for Countries boundaries from marine regions ####

###_______________________________Read Me_____________________________________________#
#### Data License and references ###
# Marine Regions’ data is licensed under CC-By-NC-SA (https://creativecommons.org/licenses/by-nc-sa/4.0/).
# Flanders Marine Institute (2018). Maritime Boundaries Geodatabase: Maritime Boundaries and Exclusive Economic Zones (200NM), version 10. Available online at http://www.marineregions.org/. https://doi.org/10.14284/312

#### Using the SAU shapefile that contains EEZ per ocean (E.G. Mexico Pacific and Mexico Atlantic)
###________________________________________________________________________________#

@@

# The path
path_world <- paste(Data_Path,"Spatial_Data/SAU_Shapefile/",sep="")
# #The File
fnam_world <- "SAUEEZ_July2015.shp"
# #Load it!
# 
World_EEZs <- st_read(dsn = path_world,
                      layer =file_path_sans_ext(fnam_world))

@@

@@

####_________________________________________ TESTING PERIOD_________________________________________####
# I'm substetting central america to do the whole analysis (smaller EEZs = faster computation) once I have figure out things I'll run for all

@@

@@

```{r Species_Status_fun, eval = T, echo =F}

@@

  Data <- GetSppDist(Spp, Model = "SAU_C", Coord = Coord) %>% 
    set_names(c(seq(1950,2014,1),"INDEX","Species","Longitude","Latitude")) %>% 
    gather("Year","Value",`1950`:`2014`) %>%
    arrange(INDEX) %>% 
    left_join(Index_Code,
              by = "INDEX") %>% 
    filter(!is.na(EEZID),
           Value > 0) %>% 
    group_by(TaxonKey=Species,Year,Territory) %>% # Total catch per country
    summarise(Total_Catch = sum(Value,na.rm=T)) %>% 
    group_by(TaxonKey,Territory,Year) %>% 
    mutate(RMean = rollmean(x = Total_Catch, 3,
                            align = "right",
                            fill = Total_Catch)
           ) %>% # Running mean
    select(-Total_Catch) %>% 
    ungroup() %>% 
    mutate(Year = as.numeric(Year))

  # Tresholds to evaluate stocks if these are not met, then soecies status = "No Status"
  Tresholds_Data <- Data %>% 
    arrange(Territory) %>% 
    group_by(TaxonKey,Territory, grp = cumsum(c(1, diff(Year) != 65))) %>% 
    group_by(TaxonKey,Territory) %>% 
    summarise(Years_appart = ifelse(max(Year)-min(Year) >= 0,"Yes","No"), # Step 1. first and last reported landings are at least 10 years apart
              Consecutive = ifelse(max(grp) >= 5,"Yes","No"), # Step 2. There are at least 5 consecutive catch years
              Catch_over = ifelse(max(RMean >= 1000),"Yes","No") # Step 3. 3) The catch in a particular area (LME) is at least 1,000 tonnes.
    ) %>% 
    gather("Step","Treshold",Years_appart:Catch_over) %>% 
    group_by(TaxonKey,Territory) %>% 
    summarise(Met_Criteria = paste(unique(Treshold),collapse =";")
              )

@@

    # Get Variables to perform status metric
    Status_Data <- Data %>% 
      anti_join(No_Status_Data, # Removes no status data
                by = c("TaxonKey","Territory")
      ) %>% 
      # Get Max catch and max catch year
      group_by(TaxonKey,Territory) %>% 
      top_n(1,RMean) %>% 
      rename(Max_Peak = RMean,
             Y_Max_Peak = Year) %>% 
      # Get Min catch after peak catch
      left_join(Data,
                by = c("TaxonKey","Territory")
                ) %>% 
      filter(Year > Y_Max_Peak) %>% 
      group_by(TaxonKey,Territory) %>% 
      top_n(-1,RMean) %>% 
      rename(Post_Max_Min = RMean,
             Y_Post_Max_Min = Year) %>% 
      left_join(Data,
                by = c("TaxonKey","Territory")
                ) %>% 
      # Final Evaluation 
      mutate(Status = ifelse(Year > Y_Post_Max_Min & Post_Max_Min < (Max_Peak*0.10) & (RMean > (Max_Peak*0.10) & RMean < (Max_Peak*0.50)),"Rebuilding",
                             ifelse((Year < Y_Max_Peak & RMean <= (Max_Peak*0.50) | Y_Max_Peak == max(Year)),"Developing",
                                    ifelse(RMean > Max_Peak*0.50,"Max Exploited",
                                           ifelse(Year > Y_Max_Peak & (RMean > (Max_Peak*0.10) & RMean < (Max_Peak*0.50)),"Over Exploited",
                                                  ifelse(Year > Y_Max_Peak & RMean < Max_Peak*0.10,"Collapsed",
                                                         "No Status")
                                           )
                                    )
                             )
      )
      ) %>% 
      filter(Year > 2004) %>% # last 10 years of data
      group_by(TaxonKey,Territory,Status) %>% 
      summarise(n_Status= n()) %>% 
      group_by(TaxonKey,Territory) %>% 
      top_n(1,n_Status) %>%
      select(-n_Status)

@@

    # If only No status data, then just return that
    if(nrow(No_Status_Data) > 0 & nrow(Status_Data) == 0){
      return(No_Status_Data)
    }else{
      return(Status_Data)
    }

@@

```{r spp_status_data eval = T, echo = F}

@@

```{r species_status_routine eval = T, echo = F}
# Needs to have trans species 
Species_Status <- bind_rows(
    mclapply(
      Species_List,
      FUN = EstSppStatus,
      Coord = CoorG  # Gabriels coordinate system for read species dist
    )
  )

# x <-Species_Status %>% 
#   group_by(Territory,Status) %>% 
#   summarise(n())

write.csv(Species_Status,
          paste(Data_Path,"Distribution_Data/Species_Status.csv",sep=""),
                row.names = F)

@@

#### Function `GetSppDist`

This function Read in species distribution from the observation data, ESM models and SAU data

@@

      # print(paste("No info for this species"))
    }

@@

#### Function `EstTransIndex`

##### NOTES:

###### General:
- For now using SAU_C as presence absence
- Grouping by species distribution within EEZ and not EEZ area

###### Tresholds:
###### Model Index, line ~479

This first treshold estimates the uncertanty of models. Goes from 0 - 1 as follows:

** NOTE THAT THE FUNCTION IS SETTED TO OUTPUT RESULTS FOR LL MODELS **

###### Distribution Index, line ~482

>>>>

1. Determine which countries have Neighbouring EEZs  
  1.1. For this we used the Sea Around Us EEZ map that discriminate by oceans and a routine (*Estimate Transboundary Species*) to identify Neighbouring polygons,
  1.2. Then we match the SAU spatial information (e.g. INDEX within EEZ and coordinates) with the Neighbouring data

2. Overlap distribution maps with Neighbours data  
  2.1. For each species we merged the Neighbours data with their known distribution (SAU distribution and catch,ESMs, and Observation)  
3. Determine transboundary if:  
  3.1. Species is present in both Neighbouring EEZs by INDEX

We create two "confidence level" based on data and area covered, i.e;

-
-

## Determine species list

This part is to determine the list of species to be analized in the project. Initially this list was limited by the species data from the DBEM, however, as we are not using the DBEM, the limit is now the SAU.

@@

## Determine Neighbourds

This is the first step of the analysis and will determine what EEZs are neighbouring. THe routine creates a final dataset that we then use to estimate transboundary species.

### Function `EstNeighbours`

@@

  # Merge them with SAU data so we have the INDEX and lat long info
  suppressWarnings(
    # warning for characters in name, no problem
  DBEM_Neighbours_Data <- Result_table %>% 
     left_join(EEZIDs_List, 
              by = "Name") %>% # Asign EEZ id to each country
    left_join(EEZ_CellID, 
              by = "EEZID") %>% # Asign EEZ id to each country
    left_join(Coor,
              by = "INDEX") %>%  # Include lat long
    rename(Country_Territory = Name)

@@

```{r Neighbours_Control_Pannel, eval = F, echo = F}

###___________________Read Me_____________________#
#### Using the SAU shapefile that contains EEZ per ocean (E.G. Mexico Pacific and Mexico Atlantic)
###_______________________________________________#

@@

# The path
path_world <- paste(Data_Path,"Spatial_Data/SAU_Shapefile/",sep="")

# The File
fnam_world <- "SAUEEZ_July2015.shp"

# Load it!
World_EEZs <- st_read(dsn = path_world,
                      layer =file_path_sans_ext(fnam_world))

@@

```{r Neighbours_Subroutine, eval = F, echo = T}

@@

####_________________________________________ TESTING PERIOD_________________________________________###
# I'm substetting central america to do the whole analysis (smaller EEZs = faster computation) once I have figure out things I'll run for all

@@

This sub routine determines the species exploitation status based on the SAU methodology. **Note:** For the current project we do not consider this method to determine the species exploitation status *per se*, but rather a catch trend that *could* indicate that a species is under/over/max exploited.

In the original method the authors analyze stocks, while we analyze at the species level within each EEZ. Also (n), the number of species, is defined as a time series of a given species, genus or family (higher and pooled groups have been excluded) for which the first and last reported landings are at least 10 years apart, for which there are at least 5 years of consecutive catches and for which the catch in a given area (LME) is at least 1000 tonnes. We removed the last one (LME) and computed all of the species that only followed the first two;

2) There are at least five years of consecutive catches

@@

This function estimates the exploitation status of a species based on catch data.

```{r funEstSppStatus, eval = T, echo =F}

@@

  # Load species' data
  Data <- GetSppDist(Spp, Model = "SAU_C", Coord = Coord) %>% 
    set_names(c(seq(1950,2014,1),"INDEX","Species","Longitude","Latitude")) %>% 
    gather("Year","Value",`1950`:`2014`) %>%
    arrange(INDEX) %>% 
    left_join(Index_Code,
              by = "INDEX") %>% 
    filter(!is.na(EEZID),
           Value > 0) %>% 
    group_by(TaxonKey=Species,Year,Territory) %>% # Estimates total catch per country
    summarise(Total_Catch = sum(Value,na.rm=T)) %>% 
    group_by(TaxonKey,Territory,Year) %>% # Running mean as Pauly et la.
    mutate(RMean = rollmean(x = Total_Catch, 3,
                            align = "right",
                            fill = Total_Catch)
    ) %>% 
    select(-Total_Catch) %>% 
    ungroup() %>% 
    mutate(Year = as.numeric(Year))

  # Tresholds to evaluate stocks if these are not met, then soecies status = "No Status"
  Tresholds_Data <- Data %>% 
    arrange(Territory) %>% 
    group_by(TaxonKey,Territory, grp = cumsum(c(1, diff(Year) != 65))) %>% 
    group_by(TaxonKey,Territory) %>% 
    summarise(Years_appart = ifelse(max(Year)-min(Year) >= 0,"Yes","No"), # Step 1. first and last reported landings are at least 10 years apart
              Consecutive = ifelse(max(grp) >= 5,"Yes","No")#, # Step 2. There are at least 5 consecutive catch years
              # Catch_over = ifelse(max(RMean >= 1000),"Yes","No") # Step 3. 3) The catch in a particular area (LME) is at least 1,000 tonnes. (CURRENTLY OFF)
    ) %>% 
    gather("Step","Treshold",Years_appart:Catch_over) %>% 
    group_by(TaxonKey,Territory) %>% 
    summarise(Met_Criteria = paste(unique(Treshold),collapse =";")
    )

@@

  # Get Variables to perform status metric
  Status_Data <- Data %>% 
    anti_join(No_Status_Data, # Removes no status data
              by = c("TaxonKey","Territory")
    ) %>% 
    # Get Max catch and max catch year
    group_by(TaxonKey,Territory) %>% 
    top_n(1,RMean) %>% 
    rename(Max_Peak = RMean,
           Y_Max_Peak = Year) %>% 
    # Get Min catch after peak catch
    left_join(Data,
              by = c("TaxonKey","Territory")
    ) %>% 
    filter(Year > Y_Max_Peak) %>% # Filter years after peak year
    group_by(TaxonKey,Territory) %>% 
    top_n(-1,RMean) %>% # find min value
    rename(Post_Max_Min = RMean,
           Y_Post_Max_Min = Year) %>% 
    left_join(Data,
              by = c("TaxonKey","Territory")
    ) %>% 
    # Final Species Status Evaluation 
    mutate(Status = ifelse(Year > Y_Post_Max_Min & Post_Max_Min < (Max_Peak*0.10) & (RMean > (Max_Peak*0.10) & RMean < (Max_Peak*0.50)),"Rebuilding",
                           ifelse((Year < Y_Max_Peak & RMean <= (Max_Peak*0.50) | Y_Max_Peak == max(Year)),"Developing",
                                  ifelse(RMean > Max_Peak*0.50,"Max Exploited",
                                         ifelse(Year > Y_Max_Peak & (RMean > (Max_Peak*0.10) & RMean < (Max_Peak*0.50)),"Over Exploited",
                                                ifelse(Year > Y_Max_Peak & RMean < Max_Peak*0.10,"Collapsed",
                                                       "No Status")
                                         )
                                  )
                           )
    )
    ) %>% 
    filter(Year > 2004) %>% # last 10 years of data
    group_by(TaxonKey,Territory,Status) %>% 
    summarise(n_Status= n()) %>% 
    group_by(TaxonKey,Territory) %>% 
    top_n(1,n_Status) %>%
    select(-n_Status)

  # Function Return

@@

  # If only No status data, then just return No Status Data
  if(nrow(No_Status_Data) > 0 & nrow(Status_Data) == 0){
    return(No_Status_Data)
  }else{
    return(Status_Data)
  }
}

@@

```{r Spp_Status_Control_Pannel, eval = T, echo = F}

@@

```{r Spp_Status_Sub_Routine, eval = T, echo = F}
# Needs to have trans species 
Species_Status <- bind_rows(
    mclapply(
      Species_List,
      FUN = EstSppStatus,
      Coord = CoorG  # Gabriels coordinate system for read species dist
    )
  )

# Save data
# write.csv(Species_Status,
#           paste(Results_Path,"Species_Status.csv",sep=""),
#                 row.names = F)

@@

This is the main step of the project. In this segment we estimate the number of transboundary species existing today in each of the worlds EEZs.

The datasets SAU_D, OCC, and ESM are used as sppecies presence while SAU_C is used as a confirmation of that. The `Model Index` is the first treshold and estimates the uncertanty of models. Goes from 0 - 1 as follows:

The `Area_Index` is the second treshold and estimates the proportion of each species distribution in each EEZ. It only take into consideration the distribution of the species within the two EEZs, rather than the complete distribution. Hence, is mirrord (e.g. if EEZ#1 is 0.40, EEZ#2 is 0.6)

** NOTE THAT THE FUNCTION IS SETTED TO OUTPUT RESULTS FOR ALL MODELS **

### Function `GetSppDist`

This function reads in species distribution from the observation data, ESM models and SAU data.

@@

@@

### Function `EstTransIndex`

This is the main function of the project, the one that actually estimates if a species is transboundary or not

```{r funEstTransIndex, eval= T ,echo = T,warning = F,message = F}
